Title

### 2018-09-20 thesis/paper/paper.tex
Changed font of predicates and constants

\section{Datalog}
Datalog is a logic query language. It is fully declarative and uses rules to derive additional knowledge from given facts. The most basic domain element in datalog is a \emph{constant}, which will be denoted as \emph{someConstant}. In the context of Wikidata, any item or property, e.g. \emph{Earth} (Q2) or \emph{head of government} (P6), could be represented as a constant. Relations in datalog are expressed using \emph{predicates} with a specified arity \(n\), written as predicate\([n]\). To stay close to Wikidata in our examples, let us take the predicate triple\([3]\) as an example. This would be used to express the \emph{item}-\emph{property}-\emph{value} relation with each position representing one part of the triple.\\
Additionally, datalog also provides \emph{variables}, which will be capitalized, like \(X\). The sets of predicates {\bf P}, constants {\bf C} and variables {\bf V} are mutually disjoint. Constants and variables are collectively known as \emph{terms} and are, together with predicates, used to form \emph{atoms}. An atom has the form predicate\((t_1, \ldots , t_n)\) with \(n\) as the arity of the predicate and \(t_1,\ldots,t_n\) terms.\\
Atoms are the building blocks of \emph{rules}, which are used to derive information in datalog. A rule is a premise that implies a conclusion, and both are conjunctions of atoms.
Let \(P_n\) denote  premise atoms and \(C_n\) the conclusion atoms. The variables used in one rule are divided into two mutually disjoint sets \(V_p\) and \(V_r\) which are the variables occurring only in its premise atoms and occurring in conclusion atoms and premise atoms, respectively. The general from of a rule is (with \(k \in \mathbb{N}^+\), \(l \in \mathbb{N}\))
\begin{equation}
\forall V_p \forall V_r.(C_1 \wedge \ldots \wedge C_k \leftarrow P_1 \wedge \ldots \wedge P_l)
\end{equation}

@@

Within the rule predicates we can differentiate between two different schemas: the \emph{extensional schema} and the \emph{intensional schema}, called EDB and IDB (extensional/intensional database). An EDB predicate may only occur in the premise of a rule, meaning it can never be derived. All predicates that appear in the conclusion at least once are IDB predicates.

\begin{example}
For the following example, taken from Wikidata, descriptive names will be used instead of the IDs to increase readability. The individual entities and properties are introduced with their ID beforehand, so the example can be retraced. The species \emph{blue whale} (Q42196) has the \emph{parent taxon} (P171) \emph{balaenoptera} (Q133320), which in turn has the parent taxon \emph{rorqual} (Q232829). The predicates used are triple\_EDB\([3]\) as the input and isInTaxon\_IDB\([2]\), which will be derived. The basic facts are:
\[\text{triple\_EDB}(\text{blue whale}, \text{parent taxon}, \text{balaenoptera})\]
\[\text{triple\_EDB}(\text{balaenoptera}, \text{parent taxon}, \text{rorqual})\]
From this we can draw conclusions. The first step is to represent the already existing facts as IDB:
\[\text{isInTaxon\_IDB}(X, Y) \leftarrow \text{triple\_EDB}(X, \text{parent taxon}, Y)\]
Finally the transitive closure of this relation can be derived with
\[\text{isInTaxon\_IDB}(X, Y) \leftarrow \text{isInTaxon\_IDB}(X, Z) \wedge \text{isInTaxon\_IDB}(Z, Y)\]
\end{example}

@@

\subsection{Representation of Wikidata as Facts}\label{subsec_representation}
To represent the content of Wikidata as facts, the following predicates are introduced:
\begin{table}[H]
\caption{EDB predicates}\label{tab_predicates}
\begin{tabular}{llp{8.5cm}}
Predicate & Positions & Description \\
\hline
statement\_EDB & \makecell{statementID \\ entityID \\ propertyID \\ value (entityID or string)} & represents basic statements \\
\hline
qualifier\_EDB & \makecell{statementID \\ propertyID \\ value} & represents qualifiers on statements \\
\hline
reference\_EDB & \makecell{statementID \\ reference hash \\ propertyID \\ value} & represents references on statements; reference hash groups the property-value pairs belonging to one reference \\
\hline
item\_EDB & entityID & marks an entity as an item \\
property\_EDB & entityID & marks an entity as a property \\
\hline
unit\_EDB & \makecell{value \\ unitID} & denotes which unit a value has \\
\hline
rank\_EDB & \makecell{statementID \\ rank} & denotes the rank a statement has
\end{tabular}
\end{table}

@@

\subsection{Task}\label{subsec_task}
Each property constraint will be modelled and evaluated as one set of rules per property per constraint, meaning each violating statement will be of a form similar to
\begin{align*}
&\text{violation\_statement}(S, E, \text{propertyConstant}, V) &\leftarrow \text{atoms determining the violation} \\
&\text{violation\_qualifier}(S, \text{propertyConstant}, V) &\leftarrow \text{atoms determining the violation} \\
&\text{violation\_reference}(S, H, \text{propertyConstant}, V) &\leftarrow \text{atoms determining the violation}
\end{align*}
where propertyConstant denotes the constrained property. The variables \(S\), \(I\), \(V\), and \(H\) are used for statement ID, entity ID, value and reference hash, respectively.

The task for each constraint is thus: Given the constraint and a constrained property, potentially with qualifiers, which rule finds all violations of the rule in a given set of triples, qualifiers and references if applicable? The given rule set should return all violations if queried with the following atoms:
\begin{align*}
&\text{violation\_statement}(S, E, P, V) \\
&\text{violation\_qualifier}(S, P, V) \\
&\text{violation\_reference}(S, H, P, V)
\end{align*}

@@

\paragraph{Rules}
The \emph{property scope} constraint requires no context information to be violated; the triple, qualifier, or reference with the constrained property is enough. This means violations can be found using these rules:
\begin{align}
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow \\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V)\\
\end{split}\\
\begin{split}
\text{violation\_qualifier}&(S, \text{propertyConstant}, V) \leftarrow \\
&\text{qualifier\_EDB}(S, \text{propertyConstant}, V)\\
\end{split}\\
\begin{split}
\text{violation\_reference}&(S, \text{propertyConstant}, V) \leftarrow \\
&\text{reference\_EDB}(S, H, \text{propertyConstant}, V)\\
\end{split}
\end{align}
Which of these rules need to be used depends on the values of the \emph{property scope} qualifier. If a concrete constrained on a property does not have \emph{as main value} as a value of the qualifier, the first rule is used. The case for \emph{as qualifier} and \emph{as reference} is analogous with the second and third rule, respectively.

@@

\paragraph{Rules}
The conflicts-with constraint is violated if another statement on the same item either has a conflicting property or a conflicting property and a conflicting value depending on specified qualifiers. It depends on a conjunction (\(O\) denoting another statement ID, \(C\) another value) and a violation for a conflicts-with constraint is thus either rule one for constraints with no conflicting values specified or one instance of rule two for each conflicting value if specified.
\begin{align}
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow\\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V) \wedge{}\\
&\text{statement\_EDB}(O, E, \text{conflictingPropertyConstant}, C)
\end{split}\\
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow\\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V) \wedge{}\\
&\text{statement\_EDB}(O, E, \text{conflictingPropertyConstant}, \text{conflictingValueConstant})
\end{split}
\end{align}

@@

Since there are only three options (and only two in use), violations can be found without inequality by finding the statement\_EDB with the constrained property whose subject is of the type that is not allowed, meaning either \emph{Wikibase item} or \emph{Wikibase property}.

@@

\paragraph{Rules}
The allowed entity types constraint is violated by a statement with the wrong entity type.  Constraints regarding the lexeme entity type can be ignored since they are not currently used. The violations depend on a simple conjunction:
\begin{align}
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow \\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V) \wedge \text{item}(I)
\end{split}\\
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow \\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V) \wedge \text{property}(I)
\end{split}
\end{align}
The first rule is created for all \emph{allowed entity type} constraints whose \emph{allowed entity type} does not contain \emph{Wikibase item}. The second rule works analogous for \emph{Wikibase property}.

@@

\paragraph{Rules}
The none of constraint simply forbids certain values for properties, thus effectively prohibiting certain statements, qualifiers, and references. The rules, depending on nothing but specific database entries, are
\begin{align}
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, \text{forbiddenValueConstant}) \leftarrow \\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, \text{forbiddenValueConstant})
\end{split}\\
\begin{split}
\text{violation\_qualifier}&(S, \text{propertyConstant}, \text{forbiddenValueConstant}) \leftarrow \\
&\text{qualifier\_EDB}(S, \text{propertyConstant}, \text{forbiddenValueConstant})
\end{split}\\
\begin{split}
\text{violation\_reference}&(S, \text{propertyConstant}, \text{forbiddenValueConstant}) \leftarrow \\
&\text{reference\_EDB}(S, H, \text{propertyConstant}, \text{forbiddenValueConstant})
\end{split}
\end{align}
Given a specified constraint, each of the three rules is created once for each \emph{forbidden value}.

@@

\subsection{Approaches}
In every approach it will be necessary to load some basic inequalities into the database. To this end the predicate unequal\_EDB\([2]\) is introduced. Since the goal is to keep the number of applications of this predicate small, an IDB predicate unequal\([2]\) has to be introduced. This leads to the rule
\begin{align}
\text{unequal}(X, Y) \leftarrow \text{unequal\_EDB}(X, Y)
\end{align}

Given a set of unequal constants \(U\), the task is to generate rules so that for all \( x, y \in U: x \ne y \Rightarrow \text{unequal}(x, y)\).

\subsubsection{Naive}\label{subsubsec_naive}
The naive approach would be to simply add all possible combinations:
\begin{align}
\text{for all } x, y \in U, x \ne y \text{ add } \text{unequal\_EDB}(x, y)
\end{align}
This can be easily improved using the inverse and adding the following rule:
\begin{align}\label{equ_inverse}
\text{unequal}(X, Y) \leftarrow \text{unequal}(Y, X)
\end{align}
Then it is sufficient to add each combination once. To achieve this assume an arbitrary total order \(<\) has been established on the set \(E\):
\begin{align}
\text{for all } x, y \in U, x < y \text{ add } \text{unequal\_EDB}(x, y)
\end{align}

@@

An unequal constant \(u \in U\) can be represented a string and be written as \(c_0 \ldots c_{l - 1}\) with \(l = len(u)\) being the length of the string. Denote the set \(C\) as the set of characters occurring in unequal strings. Then \(l_{\max} = \max\limits_{\forall u \in U}(len(u))\) is the length of the longest entity string that occurs in the given data.
Based on this a group of predicates letter\(_i\)\_EDB[8] is introduced where \(0 \equiv i \mod 7\), \(0 \leq i < l_{max} \).

For each unequal constant string \(u = c_0 \ldots c_{l - 1}\) and \(0 \equiv i \mod 7, 0 \leq i < len(u)\) an atom using predicate letter\(_i\)\_EDB is created as 
\begin{align}
\begin{split}
\text{letter}_i\text{\_EDB}(e, f(e, i), f(e, i+1), \ldots, f(e, i+6)
\end{split}
\begin{split}
f(e, i) = 
\begin{cases}
      c_i & i < len(e) \\
      \text{none} & i \geq len(e)
\end{cases}
\end{split}
\end{align}

Now that inequality between all characters and the link between entities and their characters are introduced, the inequality between two entities can be derived by creating the following rules for all predicates letter\(_i\)\_EDB (with \text{letters} short for \(\text{letter}_i\text{\_EDB}(X, A_i, \ldots, A_{i+6}) \wedge\text{letter}_i\text{\_EDB}(Y, B_i, \ldots, B_{i+6})\), \(i \leq j < i + 7\)):
\begin{align}
\text{unequal}(X, Y) \leftarrow \text{letters} \wedge \text{unequal}(A_j, B_j)
\end{align}
Note that the facts fulfilling \(\text{unequal}(A_j, B_j)\) are created by using the approach outlined in Section \ref{subsubsec_naive} for the set of characters \(C\) in all unequal constants \(U\).

@@

To save space and avoid redundancy, the following rules will not explicitly state the inequality requirement predicate. Instead, assume that for every premise of the form \(B_1 \wedge \ldots\wedge B_m \wedge I_1 \wedge I_n\) where \(B_1 .. B_m\) are arbitrary atoms, \(I_1 .. I_n\) are atoms of the form \(\text{unequal}(t_{n_1}, t_{n_2})\) with \(t_{n_1}, t_{n_2}\) terms and all variables amount \(t_{n_1}\) and \(t_{n_2}\) appear in at least one of \(B_1 \ldots B_m\), rule one is introduced for all \(n\). The rules outlined in Section \ref{subsubsec_encoding} are changed to utilise this new requirement as shown in rule two.
\begin{align}
\begin{split}
\text{req\_inequality}(t_{n_1}, t_{n_2})
 \leftarrow B_1 \wedge \ldots \wedge B_m
\end{split}\\
\begin{split}
\text{unequal}(X, Y) \leftarrow \text{req\_inequality}	(X, Y) \wedge \text{letters} \wedge \text{unequal}(A_j, B_j)
\end{split}
\end{align}

@@

\paragraph{Rules}
The \emph{distinct values} constraint requires two distinct constrained statements with the same value. 
Two constrained statements with the same value can be found via a simple conjunction of two statement\_EDB atoms with the propertyConstant as predicate and a variable \(V\) as value. Since this would mark every constrained statement as a violation because it can fulfil both atoms at the same time, inequality of the statement IDs \(S\) and \(O\) must be tested as well. The rules for qualifiers and reference are analogous.
\begin{align}
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow \\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V) \wedge{} \\
&\text{statement\_EDB}(O, X, \text{propertyConstant}, V) \wedge{} \\
&\text{unequal}(S, O)
\end{split}
\end{align}

@@

\paragraph{Rules}
To handle this constraint, inequality between all units need to be established. Then a violation can be identified by finding all statement\_EDB, qualifier\_EDB, and reference\_EDB with the constrained property and a unit unequal to all allowed units.
The allowed units constraint needs to determine inequality between all allowed units and each unit on a constrained statement. To this end the units of a statement, qualifier, or reference need to be extracted during pre-processing and their inequality established. Quantities are rendered as string during pre-processing, specifying their value, bounds, and unit.

The rules require multiple unequal-atoms for all allowed units in the set \(A\) of allowed units. Violations are thus
\begin{align}
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow \\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V) \wedge{} \\
&\text{unit}(V, U)  {\textstyle\bigwedge}_{a \in A} \text{unequal}(a, U)
\end{split}\\
\begin{split}
\text{violation\_qualifier}&(S, \text{propertyConstant}, V) \leftarrow \\
&\text{qualifier\_EDB}(S, \text{propertyConstant}, V) \wedge{} \\
&\text{unit}(V, U) {\textstyle\bigwedge}_{a \in A} \text{unequal}(a, U)
\end{split}\\
\begin{split}
\text{violation\_reference}&(S, H, \text{propertyConstant}, V) \leftarrow \\
&\text{reference\_EDB}(S, H, \text{propertyConstant}, V) \wedge{} \\
&\text{unit}(V, U) {\textstyle\bigwedge}_{a \in A} \text{unequal}(a, U)
\end{split}
\end{align}

@@

\paragraph{Rules}
To represent a violation of the \emph{allowed qualifiers} constraint in datalog, inequality between all specified allowed qualifiers and the properties of qualifiers is required. The constraint is violated if a qualifier on a constrained statement exists that is unequal to all allowed qualifiers. The notation for set inequality as introduced in Section \ref{subsec_3_allowed_units} will be used. With \(A\) the set of allowed qualifiers the rule is:
\begin{align}
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow \\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V) \wedge{} \\
&\text{qualifier\_EDB}(S, Q, O) {\textstyle\bigwedge}_{a \in A} \text{unequal}(a, Q)
\end{split}
\end{align}

@@

\paragraph{Rules}
The one-of qualifier value constraint is violated if a constrained statement has a qualifier with a value unequal to all allowed values. Inequality needs to be established between all values of qualifiers and all allowed values. With the set of allowed Values \(A\) the rules are
\begin{align}
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow \\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V) \wedge{} \\
&\text{qualifier\_EDB}(S, \text{qualifierProperty}, O) {\textstyle\bigwedge}_{a \in A} \text{unequal}(a, O)
\end{split}
\end{align}

>>>>

\section{Datalog}
Datalog is a logic query language. It is fully declarative and uses rules to derive additional knowledge from given facts. The most basic domain element in datalog is a \emph{constant}, which will be denoted as \emph{someConstant}. In the context of Wikidata, any item or property, e.g. \emph{Earth} (Q2) or \emph{head of government} (P6), could be represented as a constant. Relations in datalog are expressed using \emph{predicates} with a specified arity \(n\), written as \textsc{predicate}\([n]\). To stay close to Wikidata in our examples, let us take the predicate \textsc{triple}\([3]\) as an example. This would be used to express the \emph{item}-\emph{property}-\emph{value} relation with each position representing one part of the triple.\\
Additionally, datalog also provides \emph{variables}, which will be capitalized, like \(X\). The sets of predicates {\bf P}, constants {\bf C} and variables {\bf V} are mutually disjoint. Constants and variables are collectively known as \emph{terms} and are, together with predicates, used to form \emph{atoms}. An atom has the form \textsc{predicate}\((t_1, \ldots , t_n)\) with \(n\) as the arity of the predicate and \(t_1,\ldots,t_n\) terms.\\
Atoms are the building blocks of \emph{rules}, which are used to derive information in datalog. A rule is a premise that implies a conclusion, and both are conjunctions of atoms.
Let \(P_n\) denote  premise atoms and \(C_n\) the conclusion atoms. The variables used in one rule are divided into two mutually disjoint sets \(V_p\) and \(V_r\) which are the variables occurring only in its premise atoms and occurring in conclusion atoms and premise atoms, respectively. The general from of a rule is (with \(k \in \mathbb{N}^+\), \(l \in \mathbb{N}\))
\begin{equation}
\forall V_p \forall V_r.(C_1 \wedge \ldots \wedge C_k \leftarrow P_1 \wedge \ldots \wedge P_l)
\end{equation}

@@

Within the rule predicates we can differentiate between two different schemas: the \emph{extensional schema} and the \emph{intensional schema}, called \textsc{edb} and \textsc{idb} (extensional/intensional database). An \textsc{edb} predicate may only occur in the premise of a rule, meaning it can never be derived. All predicates that appear in the conclusion at least once are \textsc{idb} predicates.

\begin{example}
For the following example, taken from Wikidata, descriptive names will be used instead of the IDs to increase readability. The individual entities and properties are introduced with their ID beforehand, so the example can be retraced. The species \emph{blueWhale} (Q42196) has the \emph{parentTaxon} (P171) \emph{balaenoptera} (Q133320), which in turn has the parent taxon \emph{rorqual} (Q232829). The predicates used are \textsc{triple\_edb}\([3]\) as the input and \textsc{is\_in\_taxon\_idb}\([2]\), which will be derived. The basic facts are:
\[\textsc{triple\_edb}(\emph{blueWhale}, \emph{parentTaxon}, \emph{balaenoptera})\]
\[\textsc{triple\_edb}(\emph{balaenoptera}, \emph{parenTaxon}, \emph{rorqual})\]
From this we can draw conclusions. The first step is to represent the already existing facts as idb:
\[\textsc{is\_in\_taxon\_idb}(X, Y) \leftarrow \textsc{triple\_edb}(X, \emph{parentTaxon}, Y)\]
Finally the transitive closure of this relation can be derived with
\[\textsc{is\_in\_taxon\_idb}(X, Y) \leftarrow \textsc{is\_in\_taxon\_idb}(X, Z) \wedge \textsc{is\_in\_taxon\_idb}(Z, Y)\]
\end{example}

@@

\subsection{Representation of Wikidata as Facts}\label{subsec_representation}
To represent the content of Wikidata as facts, the following predicates are introduced:
\begin{table}[H]
\caption{\textsc{edb} predicates}\label{tab_predicates}
\begin{tabular}{llp{8.5cm}}
Predicate & Positions & Description \\
\hline
\textsc{statement\_edb} & \makecell{statementID \\ entityID \\ propertyID \\ value (entityID or string)} & represents basic statements \\
\hline
\textsc{qualifier\_edb} & \makecell{statementID \\ propertyID \\ value} & represents qualifiers on statements \\
\hline
\textsc{reference\_edb} & \makecell{statementID \\ reference hash \\ propertyID \\ value} & represents references on statements; reference hash groups the property-value pairs belonging to one reference \\
\hline
\textsc{item\_edb} & entityID & marks an entity as an item \\
\textsc{property\_edb} & entityID & marks an entity as a property \\
\hline
\textsc{unit\_edb} & \makecell{value \\ unitID} & denotes which unit a value has \\
\hline
\textsc{rank\_edb} & \makecell{statementID \\ rank} & denotes the rank a statement has
\end{tabular}
\end{table}

@@

\subsection{Task}\label{subsec_task}
Each property constraint will be modelled and evaluated as one set of rules per property per constraint, meaning each violating statement will be of a form similar to
\begin{align*}
&\textsc{violation\_statement}(S, E, \emph{constrainedProperty}, V) &\leftarrow \emph{atoms determining the violation} \\
&\textsc{violation\_qualifier}(S, \emph{constrainedProperty}, V) &\leftarrow \emph{atoms determining the violation} \\
&\textsc{violation\_reference}(S, H, \emph{constrainedProperty}, V) &\leftarrow \emph{atoms determining the violation}
\end{align*}
where propertyConstant denotes the constrained property. The variables \(S\), \(I\), \(V\), and \(H\) are used for statement ID, entity ID, value and reference hash, respectively.

The task for each constraint is thus: Given the constraint and a constrained property, potentially with qualifiers, which rule finds all violations of the rule in a given set of triples, qualifiers and references if applicable? The given rule set should return all violations if queried with the following atoms:
\begin{align*}
&\textsc{violation\_statement}(S, E, P, V) \\
&\textsc{violation\_qualifier}(S, P, V) \\
&\textsc{violation\_reference}(S, H, P, V)
\end{align*}

@@

\paragraph{Rules}
The \emph{property scope} constraint requires no context information to be violated; the triple, qualifier, or reference with the constrained property is enough. This means violations can be found using these rules:
\begin{align}
\begin{split}
\textsc{violation\_statement}&(S, E, \emph{constrainedProperty}, V) \leftarrow \\
&\textsc{statement\_edb}(S, E, \emph{constrainedProperty}, V)\\
\end{split}\\
\begin{split}
\textsc{violation\_qualifier}&(S, \emph{constrainedProperty}, V) \leftarrow \\
&\textsc{qualifier\_edb}(S, \emph{constrainedProperty}, V)\\
\end{split}\\
\begin{split}
\textsc{violation\_reference}&(S, \emph{constrainedProperty}, V) \leftarrow \\
&\textsc{reference\_edb}(S, H, \emph{constrainedProperty}, V)\\
\end{split}
\end{align}
Which of these rules need to be used depends on the values of the \emph{property scope} qualifier. If a concrete constrained on a property does not have \emph{as main value} as a value of the qualifier, the first rule is used. The case for \emph{as qualifier} and \emph{as reference} is analogous with the second and third rule, respectively.

@@

\paragraph{Rules}
The conflicts-with constraint is violated if another statement on the same item either has a conflicting property or a conflicting property and a conflicting value depending on specified qualifiers. It depends on a conjunction (\(O\) denoting another statement ID, \(C\) another value) and a violation for a conflicts-with constraint is thus either rule one for constraints with no conflicting values specified or one instance of rule two for each conflicting value if specified.
\begin{align}
\begin{split}
\textsc{violation\_statement}&(S, E, \emph{constrainedProperty}, V) \leftarrow\\
&\textsc{statement\_edb}(S, E, \emph{constrainedProperty}, V) \wedge{}\\
&\textsc{statement\_edb}(O, E, \emph{conflictingProperty}, C)
\end{split}\\
\begin{split}
\textsc{violation\_statement}&(S, E, \emph{constrainedProperty}, V) \leftarrow\\
&\textsc{statement\_edb}(S, E, \emph{constrainedProperty}, V) \wedge{}\\
&\textsc{statement\_edb}(O, E, \emph{conflictingProperty}, \emph{conflictingValue})
\end{split}
\end{align}

@@

Since there are only three options (and only two in use), violations can be found without inequality by finding the \textsc{statement\_edb} with the constrained property whose subject is of the type that is not allowed, meaning either \emph{Wikibase item} or \emph{Wikibase property}.

@@

\paragraph{Rules}
The allowed entity types constraint is violated by a statement with the wrong entity type.  Constraints regarding the lexeme entity type can be ignored since they are not currently used. The violations depend on a simple conjunction:
\begin{align}
\begin{split}
\textsc{violation\_statement}&(S, E, \emph{constrainedProperty}, V) \leftarrow \\
&\textsc{statement\_edb}(S, E, \emph{constrainedProperty}, V) \wedge \textsc{item\_edb}(I)
\end{split}\\
\begin{split}
\textsc{violation\_statement}&(S, E, \emph{constrainedProperty}, V) \leftarrow \\
&\textsc{statement\_edb}(S, E, \emph{constrainedProperty}, V) \wedge \textsc{property\_edb}(I)
\end{split}
\end{align}
The first rule is created for all \emph{allowed entity type} constraints whose \emph{allowed entity type} does not contain \emph{Wikibase item}. The second rule works analogous for \emph{Wikibase property}.

@@

\paragraph{Rules}
The none of constraint simply forbids certain values for properties, thus effectively prohibiting certain statements, qualifiers, and references. The rules, depending on nothing but specific database entries, are
\begin{align}
\begin{split}
\textsc{violation\_statement}&(S, E, \emph{constrainedProperty}, \emph{forbiddenValue}) \leftarrow \\
&\textsc{statement\_edb}(S, E, \emph{constrainedProperty}, \emph{forbiddenValue})
\end{split}\\
\begin{split}
\textsc{violation\_qualifier}&(S, \emph{constrainedProperty}, \emph{forbiddenValue}) \leftarrow \\
&\textsc{qualifier\_edb}(S, \emph{constrainedProperty}, \emph{forbiddenValue})
\end{split}\\
\begin{split}
\textsc{violation\_reference}&(S, \emph{constrainedProperty}, \emph{forbiddenValue}) \leftarrow \\
&\textsc{reference\_edb}(S, H, \emph{constrainedProperty}, \emph{forbiddenValue})
\end{split}
\end{align}
Given a specified constraint, each of the three rules is created once for each \emph{forbidden value}.

@@

\subsection{Approaches}
In every approach it will be necessary to load some basic inequalities into the database. To this end the predicate unequal\_edb\([2]\) is introduced. Since the goal is to keep the number of applications of this predicate small, an \emph{idb} predicate unequal\([2]\) has to be introduced. This leads to the rule
\begin{align}
\textsc{unequal}(X, Y) \leftarrow \emph{unequal\_edb}(X, Y)
\end{align}

Given a set of unequal constants \(U\), the task is to generate rules so that for all \( x, y \in U: x \ne y \Rightarrow \textsc{unequal}(x, y)\).

\subsubsection{Naive}\label{subsubsec_naive}
The naive approach would be to simply add all possible combinations:
\begin{align}
\emph{for all } x, y \in U, x \ne y \emph{ add } \emph{unequal\_edb}(x, y)
\end{align}
This can be easily improved using the inverse and adding the following rule:
\begin{align}\label{equ_inverse}
\textsc{unequal}(X, Y) \leftarrow \textsc{unequal}(Y, X)
\end{align}
Then it is sufficient to add each combination once. To achieve this assume an arbitrary total order \(<\) has been established on the set \(E\):
\begin{align}
\emph{for all } x, y \in U, x < y \emph{ add } \emph{unequal\_edb}(x, y)
\end{align}

@@

An unequal constant \(u \in U\) can be represented a string and be written as \(c_0 \ldots c_{l - 1}\) with \(l = len(u)\) being the length of the string. Denote the set \(C\) as the set of characters occurring in unequal strings. Then \(l_{\max} = \max\limits_{\forall u \in U}(len(u))\) is the length of the longest entity string that occurs in the given data.
Based on this a group of predicates letter\(_i\)\_edb[8] is introduced where \(0 \equiv i \mod 7\), \(0 \leq i < l_{max} \).

For each unequal constant string \(u = c_0 \ldots c_{l - 1}\) and \(0 \equiv i \mod 7, 0 \leq i < len(u)\) an atom using predicate \emph{letter\(_i\)\_edb} is created as 
\begin{align}
\begin{split}
\emph{letter}_i\emph{\_edb}(e, f(e, i), f(e, i+1), \ldots, f(e, i+6)
\end{split}
\begin{split}
f(e, i) = 
\begin{cases}
      c_i & i < len(e) \\
      \emph{none} & i \geq len(e)
\end{cases}
\end{split}
\end{align}

Now that inequality between all characters and the link between entities and their characters are introduced, the inequality between two entities can be derived by creating the following rules for all predicates \emph{letter\(_i\)\_edb} (with \emph{letters} short for \(\emph{letter}_i\emph{\_edb}(X, A_i, \ldots, A_{i+6}) \wedge\emph{letter}_i\emph{\_edb}(Y, B_i, \ldots, B_{i+6})\), \(i \leq j < i + 7\)):
\begin{align}
\textsc{unequal}(X, Y) \leftarrow \emph{letters} \wedge \textsc{unequal}(A_j, B_j)
\end{align}
Note that the facts fulfilling \(\textsc{unequal}(A_j, B_j)\) are created by using the approach outlined in Section \ref{subsubsec_naive} for the set of characters \(C\) in all unequal constants \(U\).

@@

To save space and avoid redundancy, the following rules will not explicitly state the inequality requirement predicate. Instead, assume that for every premise of the form \(B_1 \wedge \ldots\wedge B_m \wedge I_1 \wedge I_n\) where \(B_1 .. B_m\) are arbitrary atoms, \(I_1 .. I_n\) are atoms of the form \(\textsc{unequal}(t_{n_1}, t_{n_2})\) with \(t_{n_1}, t_{n_2}\) terms and all variables amount \(t_{n_1}\) and \(t_{n_2}\) appear in at least one of \(B_1 \ldots B_m\), rule one is introduced for all \(n\). The rules outlined in Section \ref{subsubsec_encoding} are changed to utilise this new requirement as shown in rule two.
\begin{align}
\begin{split}
\textsc{req\_inequality}(t_{n_1}, t_{n_2})
 \leftarrow B_1 \wedge \ldots \wedge B_m
\end{split}\\
\begin{split}
\textsc{unequal}(X, Y) \leftarrow \textsc{req\_inequality}	(X, Y) \wedge \emph{letters} \wedge \textsc{unequal}(A_j, B_j)
\end{split}
\end{align}

@@

\paragraph{Rules}
The \emph{distinct values} constraint requires two distinct constrained statements with the same value. 
Two constrained statements with the same value can be found via a simple conjunction of two \textsc{statement\_edb} atoms with the propertyConstant as predicate and a variable \(V\) as value. Since this would mark every constrained statement as a violation because it can fulfil both atoms at the same time, inequality of the statement IDs \(S\) and \(O\) must be tested as well. The rules for qualifiers and reference are analogous.
\begin{align}
\begin{split}
\textsc{violation\_statement}&(S, E, \emph{constrainedProperty}, V) \leftarrow \\
&\textsc{statement\_edb}(S, E, \emph{constrainedProperty}, V) \wedge{} \\
&\textsc{statement\_edb}(O, X, \emph{constrainedProperty}, V) \wedge{} \\
&\textsc{unequal}(S, O)
\end{split}
\end{align}

@@

\paragraph{Rules}
To handle this constraint, inequality between all units need to be established. Then a violation can be identified by finding all \textsc{statement\_edb}, \textsc{qualifier\_edb}, and \textsc{reference\_edb} with the constrained property and a unit unequal to all allowed units.
The allowed units constraint needs to determine inequality between all allowed units and each unit on a constrained statement. To this end the units of a statement, qualifier, or reference need to be extracted during pre-processing and their inequality established. Quantities are rendered as string during pre-processing, specifying their value, bounds, and unit.

The rules require multiple unequal-atoms for all allowed units in the set \(A\) of allowed units. Violations are thus
\begin{align}
\begin{split}
\textsc{violation\_statement}&(S, E, \emph{constrainedProperty}, V) \leftarrow \\
&\textsc{statement\_edb}(S, E, \emph{constrainedProperty}, V) \wedge{} \\
&\textsc{unit\_edb}(V, U)  {\textstyle\bigwedge}_{a \in A} \textsc{unequal}(a, U)
\end{split}\\
\begin{split}
\textsc{violation\_qualifier}&(S, \emph{constrainedProperty}, V) \leftarrow \\
&\textsc{qualifier\_edb}(S, \emph{constrainedProperty}, V) \wedge{} \\
&\textsc{unit\_edb}(V, U) {\textstyle\bigwedge}_{a \in A} \textsc{unequal}(a, U)
\end{split}\\
\begin{split}
\textsc{violation\_reference}&(S, H, \emph{constrainedProperty}, V) \leftarrow \\
&\textsc{reference\_edb}(S, H, \emph{constrainedProperty}, V) \wedge{} \\
&\textsc{unit\_edb}(V, U) {\textstyle\bigwedge}_{a \in A} \textsc{unequal}(a, U)
\end{split}
\end{align}

@@

\paragraph{Rules}
To represent a violation of the \emph{allowed qualifiers} constraint in datalog, inequality between all specified allowed qualifiers and the properties of qualifiers is required. The constraint is violated if a qualifier on a constrained statement exists that is unequal to all allowed qualifiers. The notation for set inequality as introduced in Section \ref{subsec_3_allowed_units} will be used. With \(A\) the set of allowed qualifiers the rule is:
\begin{align}
\begin{split}
\textsc{violation\_statement}&(S, E, \emph{constrainedProperty}, V) \leftarrow \\
&\textsc{statement\_edb}(S, E, \emph{constrainedProperty}, V) \wedge{} \\
&\textsc{qualifier\_edb}(S, Q, O) {\textstyle\bigwedge}_{a \in A} \textsc{unequal}(a, Q)
\end{split}
\end{align}

@@

\paragraph{Rules}
The one-of qualifier value constraint is violated if a constrained statement has a qualifier with a value unequal to all allowed values. Inequality needs to be established between all values of qualifiers and all allowed values. With the set of allowed Values \(A\) the rules are
\begin{align}
\begin{split}
\textsc{violation\_statement}&(S, E, \emph{constrainedProperty}, V) \leftarrow \\
&\textsc{statement\_edb}(S, E, \emph{constrainedProperty}, V) \wedge{} \\
&\textsc{qualifier\_edb}(S, \emph{qualifierProperty}, O) {\textstyle\bigwedge}_{a \in A} \textsc{unequal}(a, O)
\end{split}
\end{align}
